---
title: "The Bank Tutorial: adapted for ICU model"
author: "Duncan Garmonsway - truncated by Tom Lawton"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
---

```{r, cache = FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>",
                      fig.width = 6, fig.height = 4, fig.align = "center")
```

## Introduction

This tutorial is a truncated version of the r-simmer vignette "The Bank Tutorial", focussing specifically on the parts required to build the basic ICU model.

## A single customer

In this tutorial we model a simple bank with customers arriving at random. We
develop the model step-by-step, starting out simply, and producing a running
program at each stage.

A simulation should always be developed to answer a specific question; in these
models we investigate how changing the number of bank servers or tellers might
affect the waiting time for customers.

### A customer arriving at a fixed time

We first model a single customer who arrives at the bank for a visit, looks
around at the decor for a time and then leaves. There is no queueing. First we
will assume his arrival time and the time he spends in the bank are fixed.

The arrival time is fixed at 5, and the time spent in the bank is fixed at 10.
We interpret '5' and '10' as '5 minutes' and '10 minutes'.  The simulation runs
for a maximum of 100 minutes, or until all the customers that are generated
complete their trajectories.

Note where these constants appear in the code below.

```{r}
library(simmer)

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  timeout(10) %>%
  log_("I must leave")

bank <-
  simmer("bank") %>%
  add_generator("Customer", customer, at(5))

bank %>% run(until = 100)
bank %>% get_mon_arrivals()
```

The short trace printed out by the `get_mon_arrivals` function shows the result.
The program finishes at simulation time 15 because there are no further events
to be executed. At the end of the visit, the customer has no more actions and no
other objects or customers are active.

### A customer arriving at random

Now we extend the model to allow our customer to arrive at a random simulated
time though we will keep the time in the bank at 10, as before.

The change occurs in the arguments to the `add_generator` function. The function
`rexp` draws from an exponential distribution with the given parameter, which in
this case is `1/5`.  See `?rexp` for more details.  We also seed the random
number generator with 10211 so that the same sequence of random numbers will be
drawn every time the script is run.

```{r}
library(simmer)

set.seed(10212)

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  timeout(10) %>%
  log_("I must leave")

bank <-
  simmer("bank") %>%
  add_generator("Customer", customer, at(rexp(1, 1/5)))

bank %>% run(until = 100)
bank %>% get_mon_arrivals()
```

The trace shows that the customer now arrives at time 7.839305. Changing the
seed value would change that time.

### Many customers

Another change will allow us to have more customers. To make things clearer we
do not use random numbers in this model.

The change is in the `add_generator` function, where we use a convenience
function `from_to` to create a sequence of start times for five customers,
starting at time 0, with an interarrival time of 10 between each customer.  One
idiosyncracy of the syntax is that no arrival is created on the `to` time, so we
give it as 41, one unit after the last arrival to be generated.  Another is that
the interarrival time must be specified as a function, hence we define a
constant function `function() {10}`

```{r}
library(simmer)

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  timeout(12) %>%
  log_("I must leave")

bank <-
  simmer("bank") %>%
  add_generator("Customer", customer, from_to(0, 41, function() {10}))

bank %>% run(until = 400)
bank %>% get_mon_arrivals()
```

### Many random customers

We now extend this model to allow arrivals at random. In simulation this is
usually interpreted as meaning that the times between customer arrivals are
distributed as exponential random variates. There is little change in our
program.  The only difference between this and the previous example of a single
customer generated at a random time is that this example generates several
customers at different random times.

The change occurs in the arguments to the `add_generator` function. The function
`rexp` draws from an exponential distribution with the given parameter, which in
this case is `1/10`.  See `?rexp` for more details.  We also seed the random
number generator with 1289 so that the same sequence of random numbers will be
drawn every time the script is run.  The 0 is the time of the first customer,
then four random interarrival times are drawn, and a final -1 stops the
generator.

The reason why we cannot use the `from_to` function here is that we want to
control the number of arrivals that are generated, rather than the end-time of
arrival generation.

```{r}
library(simmer)

set.seed(1289)

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  timeout(12) %>%
  log_("I must leave")

bank <-
  simmer("bank") %>%
  add_generator("Customer", customer, function() {c(0, rexp(4, 1/10), -1)})

bank %>% run(until = 400)
bank %>% get_mon_arrivals()
```

## A Service counter

So far, the model has been more like an art gallery, the customers entering,
looking around, and leaving. Now they are going to require service from the bank
clerk. We extend the model to include a service counter that will be modelled as
a 'resource'. The actions of a Resource are simple: a customer requests a unit
of the resource (a clerk).  If one is free, then the customer gets service (and
the unit is no longer available to other customers). If there is no free clerk,
then the customer joins the queue (managed by the resource object) until it is
the customer's turn to be served. As each customer completes service and
releases the unit, the clerk can start serving the next in line.

### One Service counter

The service counter is created with the `add_resource` function. Default
arguments specify that it can serve one customer at a time, and has infinite
queueing capacity.

The `seize` function causes the customer to join the queue at the counter.  If
the queue is empty and the counter is available (not serving any customers),
then the customer claims the counter for itself and moves onto the `timeout`
step.  Otherwise the customer must wait until the counter becomes available.
Behaviour of the customer while in the queue is controlled by the arguments of
the `seize` function, rather than by any other functions.  Once the `timeout`
step is complete, the `release` function causes the customer to make the counter
available to other customers in the queue.

Since the activity trace does not produce the waiting time by default, this is
calculated and appended using the `transform` function.

```{r, message = FALSE}
library(simmer)

set.seed(1234)

bank <- simmer()

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  set_attribute("start_time", function() {now(bank)}) %>%
  seize("counter") %>%
  log_(function() {paste("Waited: ", now(bank) - get_attribute(bank, "start_time"))}) %>%
  timeout(12) %>%
  release("counter") %>%
  log_(function() {paste("Finished: ", now(bank))})

bank <-
  simmer("bank") %>%
  add_resource("counter") %>%
  add_generator("Customer", customer, function() {c(0, rexp(4, 1/10), -1)})

bank %>% run(until = 400)
bank %>%
  get_mon_arrivals() %>%
  transform(waiting_time = end_time - start_time - activity_time)
```

Examining the trace we see that the first two customers get instant service but
the others have to wait. We still only have five customers, so we cannot draw
general conclusions.

### A server with a random service time

This is a simple change to the model in that we retain the single service
counter but make the customer service time a random variable. As is traditional
in the study of simple queues we first assume an exponential service time.

Note that the argument to `timeout` must be a function, otherwise it would apply
a constant timeout to every customer.

```{r, message = FALSE}
library(simmer)

set.seed(1269)

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  set_attribute("start_time", function() {now(bank)}) %>%
  seize("counter") %>%
  log_(function() {paste("Waited: ", now(bank) - get_attribute(bank, "start_time"))}) %>%
  # timeout(rexp(1, 1/12)) would generate a single random time and use it for
  # every arrival, whereas the following line generates a random time for each
  # arrival
  timeout(function() {rexp(1, 1/12)}) %>%
  release("counter") %>%
  log_(function() {paste("Finished: ", now(bank))})

bank <-
  simmer("bank") %>%
  add_resource("counter") %>%
  add_generator("Customer", customer, function() {c(0, rexp(4, 1/10), -1)})

bank %>% run(until = 400)
bank %>%
  get_mon_arrivals() %>%
  transform(waiting_time = end_time - start_time - activity_time)
```

This model with random arrivals and exponential service times is an example of
an M/M/1 queue and could rather easily be solved analytically to calculate the
steady-state mean waiting time and other operating characteristics. (But not so
easily solved for its transient behavior.)


### The bank with a monitor (aka summary statistics)

We now demonstrate how to calculate average waiting times for our customers.  In
the original SimPy version of this tutorial, this involved using 'Monitors'.  In
simmer, data is returned by the `get_mon_*` family of functions, as has already
been demonstrated.  Here, we simply summarise the data frame returned by the
`get_mon_arrivals` function, using standard R functions.

We also increase the number of customers to 50 (find the number '49' in the
code.code).

```{r, message = FALSE}
library(simmer)

set.seed(100005)

customer <-
  trajectory("Customer's path") %>%
  seize("counter") %>%
  timeout(function() {rexp(1, 1/12)}) %>%
  release("counter")

bank <-
  simmer("bank") %>%
  add_resource("counter", 2) %>%
  add_generator("Customer", customer, function() {c(0, rexp(49, 1/10), -1)})

bank %>% run(until = 1000)

result <-
  bank %>%
  get_mon_arrivals() %>%
  transform(waiting_time = end_time - start_time - activity_time)
```

The average waiting time for 50 customers in this 2-counter system is more
reliable (i.e., less subject to random simulation effects) than the times we
measured before but it is still not sufficiently reliable for real-world
decisions. We should also replicate the runs using different random number
seeds. The result of this run is:

```{r, message = FALSE}
paste("Average wait for ", sum(result$finished), " completions was ",
      mean(result$waiting_time), "minutes.")
```

### Multiple runs

To get a number of independent measurements we must replicate the runs using
different random number seeds. Each replication must be independent of previous
ones, so the environment (bank) must be redefined for each
run, so that the random interarrival times in the `add_generator` function are
generated from scratch.

We take the chunks of code that build the environment (bank) and run the
simulation, and wrap them in the `mclapply` function from the 'parallel'
package.  This function runs each simulation in parallel, using the available
cores in the computer's processor.  Because we use seeds for reproducability, we
pass these to the function that runs the simulation (`function(the_seed)`).

```{r, message = FALSE}
library(simmer)
library(parallel)

customer <-
  trajectory("Customer's path") %>%
  seize("counter") %>%
  timeout(function() {rexp(1, 1/12)}) %>%
  release("counter")

mclapply(c(393943, 100005, 777999555, 319999772), function(the_seed) {
  set.seed(the_seed)

  bank <-
    simmer("bank") %>%
    add_resource("counter", 2) %>%
    add_generator("Customer", customer, function() {c(0, rexp(49, 1/10), -1)})

  bank %>% run(until = 400)
  result <-
    bank %>%
    get_mon_arrivals() %>%
    transform(waiting_time = end_time - start_time - activity_time)
  paste("Average wait for ", sum(result$finished), " completions was ",
        mean(result$waiting_time), "minutes.")
}) %>% unlist()
```

The results show some variation. Remember, though, that the system is still only
operating for 50 customers, so the system may not be in steady-state.

## Balking and reneging customers

Balking occurs when a customer refuses to join a queue if it is too long.
Reneging (or, better, abandonment) occurs if an impatient customer gives up
while still waiting and before being served.

### Balking customers

Another term for a system with balking customers is one where “blocked
customers” are “cleared”, termed by engineers a BCC system. This is very
convenient analytically in queueing theory and formulae developed using this
assumption are used extensively for planning communication systems. The easiest
case is when no queueing is allowed.

As an example let us investigate a BCC system with a single server but the
waiting space is limited. We will estimate the rate of balking when the maximum
number in the queue is set to 1. On arrival into the system the customer must
first check to see if there is room. If there is not enough room, the customer
balks.

To get the balking rate, we first count the number of arrivals that didn't
finish, using the data given by `get_mon_arrivals()`.  Then we divide it by the
current model time from `now(bank)`.

```{r}
library(simmer)

timeInBank <- 12 # mean, minutes
ARRint <- 10     # mean, minutes
numServers <- 1  # servers
maxInSystem <- 2 # customers
maxInQueue <- maxInSystem - numServers

maxNumber <- 8
maxTime <- 400  # minutes
set.seed(59098)

bank <- simmer()

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  set_attribute("start_time", function() {now(bank)}) %>%
  seize("counter",
        continue = FALSE,
        reject =
          trajectory("Balked customer") %>%
          log_("BALKING")) %>%
  log_(function() {paste("Waited", now(bank) - get_attribute(bank, "start_time"))}) %>%
  timeout(function() {rexp(1, 1/timeInBank)}) %>%
  release("counter") %>%
  log_("Finished")

bank <-
  simmer("bank") %>%
  add_resource("counter",
               capacity = numServers,
               queue_size = maxInQueue) %>%
  add_generator("Customer",
                customer,
                at(c(0, cumsum(rexp(maxNumber - 1, 1 / ARRint)))))

bank %>% run(until = maxTime)

number_balked <- sum(!get_mon_arrivals(bank)$finished)
paste("Balking rate is", number_balked / now(bank), "customers per minute.")
```

When Customer2 arrives, Customer0 is already in service and Customer1 is
waiting.  There is no room, so Customer2 balks. By the vagaries of exponential
random numbers, Customer0 takes a very long time to serve (22.7358 minutes) so
the first one to find room is number Customer6 at 25.5339.

### Reneging (or abandoning) customers

Often in practice an impatient customer will leave the queue before being
served. Simmer can model this reneging behaviour using the `renege_in()`
function in a trajectory.  This defines the maximum time that a customer will
wait before reneging, as well as an 'out' trajectory for them to follow when
they renege.

If the customer reaches the server before reneging, then their impatience must
be cancelled with the `renege_abort()` function.

```{r, message = FALSE}
library(simmer)

timeInBank <- 15 # mean, minutes
ARRint <- 10     # mean, minutes
numServers <- 1  # servers

maxNumber <- 5
maxTime <- 400   # minutes
maxWaitTime <- 12 # minutes, maximum time to wait before reneging
set.seed(59030)

bank <- simmer()

customer <-
  trajectory("Customer's path") %>%
  log_("Here I am") %>%
  set_attribute("start_time", function() {now(bank)}) %>%
  renege_in(maxWaitTime,
            out = trajectory("Reneging customer") %>%
              log_(function() {
                paste("Waited", now(bank) - get_attribute(bank, "start_time"), "I am off")
              })) %>%
  seize("counter") %>%
  renege_abort() %>% # Stay if I'm being attended within maxWaitTime
  log_(function() {paste("Waited", now(bank) - get_attribute(bank, "start_time"))}) %>%
  timeout(function() {rexp(1, 1/timeInBank)}) %>%
  release("counter") %>%
  log_("Completed")

bank <-
  simmer("bank") %>%
  add_resource("counter",
               capacity = numServers) %>%
  add_generator("Customer",
                customer,
                at(c(0, cumsum(rexp(maxNumber - 1, 1 / ARRint)))))

bank %>% run(until = maxTime)
```

Customer1 arrives after Customer0 but has only 12 minutes patience. After that
time in the queue (at time 28.5058) he abandons the queue to leave Customer2 to
take his place.  Customer2 and Customer3 also renege. Customer4 is served within
12 minutes.
